Initial environmental impacts of Piney Point wastewater discharge into Tampa Bay, Florida

### 2021-08-16 manu-draft.Rmd
pca and correlation figure added

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to preven impacts to the environment.  The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established in 2001 located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FLDEP).  Decreasing holding capacity of the ponds with seasonal rain events and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.  Discharges of wastewater from the stacks occurred in 2010 to nearby Bishops Harbor connected to Tampa Bay [@Garrett11].  Recently, FLDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.

@@

* Figure \@ref(fig:map) map
* Figure \@ref(fig:timeline) timeline of events
* Figure \@ref(fig:wqmap) map of water quality observations
* Figure \@ref(fig:wqtrnds) boxplots by week with +/-1sd for monthly long-term
* Fig x seagrass and macroalgae, correlation matrix
* Fig x Red tide historic and current, include fish kill reports

@@

>>>>

The geology of central Florida is rich in phosphates that have supported a multi-billion dollar mining industry for fertilizer used in food production [@Henderson04].  By 2001, an estimated 40 million tons of phosphogypsum were created each year in northern and central Florida [@Burnett01].  Currently, seventeen phosphogypsum stacks (two active, five inactive, ten closed, [Florida Department of Environmental Protection](https://geodata.dep.state.fl.us/datasets/6277c3b1eeae4a818f8683fc29e6b35b_0/about)) exist in the Tampa Bay watershed with no long-term plan for closure or disposal of waste to prevent impacts to the environment.  The Piney Point facility located in Palmetto, Florida is a large phosphogypsum stack established in 2001 located less than two miles from the shore of Tampa Bay and near two aquatic preserves [@Henderson04].  Bankruptcy of the mining company responsible for the stack in 1999 transferred ownership to a third-party, with oversight by the Florida Departmental of Environmental Protection (FLDEP).  Decreasing holding capacity of the ponds with seasonal rain events and storage of dredging material from nearby Port Manatee have contributed to degradation of the facility.  Discharges of wastewater from the stacks occurred in 2010 to nearby Bishops Harbor connected to Tampa Bay [@Garrett11].  Recently, FLDEP authorized an [emergency order](https://floridadep.gov/sites/default/files/21-0323.pdf) on March 30th, 2021 to release wastewater from the stacks directly into lower Tampa Bay to prevent catastrophic failure of the berms supporting the holding ponds.

@@

* Figure \@ref(fig:map) map
* Figure \@ref(fig:timeline) timeline of events
* Figure \@ref(fig:wqmap) map of water quality observations
* Figure \@ref(fig:wqtrnds) boxplots by week with +/-1sd for monthly long-term
* Fig x pca of all wq variables, by area? 
* Fig x seagrass and macroalgae, correlation matrix
* Fig x Red tide historic and current, include fish kill reports

@@

```{r pcacors, fig.cap = 'Principal components analyses (PCA) and correlation matrices for water quality variables by week for April to July 2021 in response to wastewater discharge form Piney Point for (a) area 1, (b) area 2, and (c) area 3 (Figure \\@ref(fig:map)a).  All variables were standardized to zero mean and unit variance prior to PCA. Variables with log-normal distribution were log-transformed prior to analysis.  Pearson correlations indicate the linear strength of assocation between pairs of variables. ** p < 0.005, * p < 0.05, blank is not signifcant at $\\alpha$ = 0.05.'}
knitr::include_graphics(here('figs/pcacors.jpeg'))
```
